Supplementary Information

### 2016-12-04 SupplementaryInformation.Rmd
Removed the header block of `Supplemental Material`.
Changed heading types for Acknowledgements & references in main paper.

---
title: "Supplementary Information"
author: "Simon Goring"
date: "November 29, 2016"
output: word_document
---

@@

To overcome this challenge, we obtained a modified estimate of degrees of freedom (df) that accounts for spatial autocorrelation based on Moran's I from `modified.ttest` for each of the two spatial processes. To then produce a difference of the means that is not pairwise, we obtain the corrected df from `modified.ttest`, and recalculate the p-value for the t-test using the t estimate from the `t.test`. The corresponding code is available in [`confounding_table.R`](https://github.com/PalEON-Project/Composition_Climate/blob/master/R/confounding_table.R).

@@

# Supplemental Tables
**Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns (Gain, Loss, Presence) are expressed as percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (units: 10^2^ km^2^) occupied by the tree taxon for the pre-settlement or modern era. Taxa are sorted by the percent area lost (Column 3).  In this analysis, all PLS and FIA grid cells were used and a taxon was assumed have been gained/lost for a given grid cell if that grid cell had data in one time period but not the other.

@@

<!-- COM: scientific names needed in ST3 -->

**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al. 2010, Baker et al. 1985).  All entries in this table are identical to those in Table 1, indicating that the assessments of interaction strength are insensitive to this adjustment to mid-19th-century climates. See Supplemental Figure 2 for further details about this analysis.

@@

@@

<!-- COM: ST4: scientific names needed -->
<!-- COM: ST4: Correct column names and formatting for climate variables.  e.g. ppt should be PAnn -->
<!-- COM: Correct formatting for variable names in text.  I have tried to do this where I have seen them, but may have missed some variables.   -->
<!-- COM: For TMax, why does the sign of the vegetation change flip from negative to positive between Table 1 and this Supp Table 4?  I can understand why the sign of the climate change would alter as we play around with different climate normals, but the sign of the vegetation change should be stable.  The other climate variables seem to be unaffected.  This may be the bug you mentioned by email.   -->
**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1979 instead of 1990-2014. The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged. For ($T_{max}$), the sign of the climate change flips from negative (cooling) to positive (warming), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.  (Note that directionality of ($T_{min}$) is unchanged, consistent with 20^th^- and 21^st^-century warming).   See also Supplemental Figures 3, 4.

@@

@@

# Supplemental Figures
**Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Table 2.

@@

**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e. a climate normal closer to the time of the PLS surveys).  No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate 19th-century warming, with estimates of  warming at Fort Leavenworth, Kansas of 0.10oC decade^-1^ in the winter and 0.07oC decade^-1^ in the summer (Burnette et al. 2010).  Given this and the ~50 year offset between PLS observations and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35^o^C for ($T_{max}$), +0.5^o^C for ($T_{min}$), and +0.15^o^C for t_diff. This would imply that the reported regional increase in ($T_{min}$) between historical to modern periods (3.3^o^C, Fig. 1) should be further increased to 3.8^o^C, the reported regional decrease in ($T_{max}$)  (-0.8^o^C) should be reduced to -0.45^o^C), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C). Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for ($T_{min}$) and ($T_{diff}$) and too large for ($T_{max}$).  The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig. 4, suggesting that the findings of this paper are largely insensitive to this adjustment to account for 19th-century rises in temperature.

>>>>

---
output:
  word_document:
    reference_docx: word_template.docx
---

@@

To overcome this challenge, we obtained a modified estimate of degrees of freedom ($df$) that accounts for spatial autocorrelation based on Moran's I from `modified.ttest` for each of the two spatial processes. To then produce a difference of the means that is not pairwise, we obtain the corrected df from `modified.ttest`, and recalculate the p-value for the t-test using the $t$ estimate from the `t.test`. The corresponding code is available in [`confounding_table.R`](https://github.com/PalEON-Project/Composition_Climate/blob/master/R/confounding_table.R).

@@

# Supplemental Tables

**Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns (Gain, Loss, Presence) are expressed as percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (units: 10^2^ km^2^) occupied by the tree taxon for the pre-settlement or modern era. Taxa are sorted by the percent area lost (Column 3).  In this analysis, all PLS and FIA grid cells were used and a taxon was assumed have been gained/lost for a given grid cell if that grid cell had data in one time period but not the other.

@@

**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported from early meteorological time series at observing stations and US military forts (Burnette et al. 2010, Baker et al. 1985).  All entries in this table are identical to those in Table 1, indicating that the assessments of interaction strength are insensitive to this adjustment to mid-19th-century climates. See Supplemental Figure 2 for further details about this analysis.

@@

conf_pls$taxa <- paste0("*", sci_names, "*")
colnames(conf_pls) <- c("Taxon", "$P_{ann}$", "$T_{max}$", "$T_{diff}$", "$T_{min}$")

@@

**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1979 instead of 1990-2014. The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged. For ($T_{max}$), the sign of the climate change flips from negative (cooling, $-_c$) to non-significant ($._c$), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.  (Note that directionality of ($T_{min}$) is unchanged, consistent with 20^th^- and 21^st^-century warming).   See also Supplemental Figures 3, 4.

@@

conf_fia$taxa <- paste0("*", sci_names, "*")
colnames(conf_fia) <- c("Taxon", "$P_{ann}$", "$T_{max}$", "$T_{diff}$", "$T_{min}$")

@@

# Supplemental Figures

**Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Table 2.

@@

**Supplemental Figure 2**:  As Figure 4 in the main text, but with an adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e. a climate normal closer to the time of the PLS surveys).  No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate 19th-century warming, with estimates of  warming at Fort Leavenworth, Kansas of 0.10oC decade^-1^ in the winter and 0.07oC decade^-1^ in the summer (Burnette et al. 2010).  Given this and the ~50 year offset between PLS observations and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35^o^C for ($T_{max}$), +0.5^o^C for ($T_{min}$), and +0.15^o^C for t_diff. This would imply that the reported regional increase in ($T_{min}$) between historical to modern periods (3.3^o^C, Fig. 1) should be further increased to 3.8^o^C, the reported regional decrease in ($T_{max}$)  (-0.8^o^C) should be reduced to -0.45^o^C), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C). Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for ($T_{min}$) and ($T_{diff}$) and too large for ($T_{max}$).  The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig. 4, suggesting that the findings of this paper are largely insensitive to this adjustment to account for 19th-century rises in temperature.
